\chapter{Literature Review and Theoretical Background} 

\label{chap:litreview} 

  

\section{Core Concepts and Definitions} 

\label{sec:concepts} 

 

This section defines the foundational terminology used throughout the thesis.  

Each term is defined once and used consistently in subsequent chapters. 

  

\subsection{Project Management Office (PMO)} 

A Project Management Office (PMO) is a central part of an organization that is in charge of setting and keeping project management standards, governance frameworks and reporting structures across the whole company  \cite{pmi2021}. According to the Project Management Institute, PMOs can be classified into three types based on their degree of control and influence: 

  

\begin{itemize} 

    \item \textbf{Supportive PMO} --- provides templates, best practices, and  

    guidance with minimal direct control over projects; 

    \item \textbf{Controlling PMO} --- enforces compliance with project management  

    frameworks and requires the use of specific methodologies; 

    \item \textbf{Directive PMO} --- takes direct control of projects, with project  

    managers reporting directly to the PMO. 

\end{itemize} 

  

\subsection{Project Management Information System (PMIS)} 

A project management information system (PMIS) is a software-based system or set of tools that helps with planning, doing, tracking, and reporting on projects in a PMO setting. \cite{laudon2015}.  

Some common PMIS tools are Microsoft Project, Jira, SAP, and Power BI. A PMIS can be one big platform or a group of smaller tools that work together, which makes it hard to integrate information systems. 

\subsection{Information System Integration} 

Integrating information systems means connecting two or more software systems so  

that data can move between them automatically without any help from people \cite{hohpe2004}. In PMO settings, project tracking tools, financial systems, and reporting  

platforms usually work together. Integration can be as simple as manually putting together data or as more complex as using Application Programming Interfaces (APIs) or unified platform ecosystems to automatically sync data in real time.  

\subsection{Integration Maturity Model} 

A maturity model is a structured framework that shows how a capability 

 develops over a set of levels. \cite{paulk1993}.  

A maturity model indicates the progression an organization achieves in  

information system integration, evolving from elementary, manual practices to  

comprehensive, automated, platform-level integration. This thesis modifies  the idea 

 of maturity modeling applied to the specific situation of PMO information  systems,  

implying a four-level Integration Maturity Model as the fundamental  theoretical framework. 

  

\subsection{Business Intelligence and Dashboard Systems} 

Business Intelligence (BI) stands for the tools, methods, and ways of doing things   

used to gather, combine, analyze, and show business data to help making decisions. 

\cite{negash2004} In PMO settings, BI systems usually  show up as centralized dashboards  

that bring together data from many different projects and financial systems into one view.  

In this thesis, the presence or absence of centralized BI systems is used as  

an indicator of governance standardization maturity. 

  

 \subsection{Data Quality} 

Data quality refers to the degree to which data is accurate, complete, consistent, and timely for 

 its intended purpose. \cite{dama2017} 

When it comes to PMO reporting, the level of system integration has a direct influence 

 on the quality of the data: Data that is put together manually by humans is more likely 

 to have errors and require more time, while automated integration lowers the chance of 

 inconsistencies. This thesis assesses data quality via perceived accuracy and timeliness as 

 indicated by PMO representatives. 

  

 

\subsection{Reporting Automation} 

Reporting automation is the use of software tools and system integrations  

to automatically generate project status reports, dashboards, and portfolio  

summaries without manual data entry \cite{pmi2021}. Automated reporting  

reduces the workload on PMO staff and improves the timeliness and consistency  

of management information. In this thesis, reporting automation refers to  

the degree to which PMO reports are produced automatically from integrated  

systems rather than compiled manually. 

 

 

\subsection{Medium and Large Enterprises in the Latvian Context} 

As per the European Commission's classification for company sizes, A medium-sized company 

 is one that has between 50 and 249 employees and A large company employs 250 or more 

 people, \cite{ec2003}. This thesis focuses on medium and large Latvian enterprises due to their 

 higher tendency to have set up formal PMO structures and information that is complex enough 

 system landscapes to make integration maturity a real issue  

\section{Related Work: Themes, Methods, and Evidence} 

\label{sec:related} 

This section surveys existing research on PMO digitalization, information system  

integration, and maturity models. Prior work is organized into five thematic  

strands, which are contrasted to identify the gap that the present thesis addresses. 

 

 

\subsection{Traditional PMO Research and Manual Integration Practices} 

Early research on Project Management Offices focused primarily on process  

standardization, governance frameworks, and the role of PMOs in improving  

project delivery outcomes \cite{hobbs2007}. In this body of literature,  

information systems were treated as secondary tools — typically spreadsheets  

and standalone reporting software —rather than integrated platforms. Manual  

data consolidation was the dominant practice, with PMO staff collecting data  

from multiple sources and compiling reports manually. While this approach  

established the theoretical foundations of PMO governance, it did not address  

the challenges of system fragmentation or the consequences of low integration  

maturity on reporting efficiency and data quality. The primary limitation of  

this strand is its failure to account for the growing complexity of modern  

PMO information system landscapes. 

 

\subsection{API-Based and Platform-Level Integration Research} 

A second line of research looks at the technical aspects of information   

 integration of systems, with a focus on API-based connectivity and platform-level   

 ecosystems \cite{hohpe2004}. Research in this domain illustrates that organizations   

 adopting API-driven integration architectures attains significantly higher levels   

 of reporting automation and data consistency compared to those that depend on manual   

 or file-based data transfer\cite{laudon2015}. Ecosystems for platforms like   

 Native integration features in Microsoft 365, SAP, and Jira cut down on the need for manual  

work. This research strand is primarily focused on technical aspects and does not methodically  

analyze the organizational or governance effects of integration maturity in PMO-specific situations.  Furthermore, implementation costs and the technical expertise required for  

API-based integration present significant barriers for medium-sized enterprises. 

 

\subsection{Business Intelligence and Governance Standardization} 

A third research strand examines the correlation between Business Intelligence (BI) systems and organizational governance \cite{negash2004}. Studies consistently   

show that enterprises with centralized BI dashboards have higher levels   

 of governance standardization, as real-time data visibility encourages better, consistent  

decision-making and performance monitoring across projects\cite{pmi2021}. This  

strand is particularly relevant to the present thesis, as it provides theoretical  

support for Hypothesis 4 — that the use of centralized BI systems is associated  

with higher governance standardization in PMO processes. The main limitation of  

this research strand is its focus on large enterprises, with limited attention  

to medium-sized organizations or specific regional contexts. 

 

\subsection{Maturity Model Frameworks in Information Systems Research} 

The fourth strand covers research on maturity models used in information systems  

and organizational capabilities \cite{paulk1993}. Well-known frameworks such as  

CMMI, OPM3, and P3M3 help organizations measure and improve their capabilities  

across defined levels. More recent work has extended these ideas to digital  

transformation, producing Digital Maturity Models that assess how digital an  

organization is across multiple dimensions \cite{gartner2020}. However, none  

of these frameworks specifically address IS integration maturity in PMO  

environments, which is the gap this thesis fills. 

 

\subsection{Digital Maturity Research in Baltic and Eastern European Enterprises} 

The fifth strand comprises new studies on digital transformation and   

 maturity in enterprises in the Baltic and Eastern Europe. Research demonstrates that   

 companies in Latvia and the rest of the Baltic region tend to have lower levels of   

 digital maturity compared to Nordic peers, especially in the areas of integrating systems 

 and automating processes\cite{eurostat2023}. While some  

research has examined e-government digitalization and SME digital adoption in  

Latvia, empirical studies specifically addressing information system integration  

maturity in PMO environments remain absent from the literature. This represents  

a significant gap, as the Latvian enterprise context — characterized by a  

predominance of medium-sized organizations and a rapidly evolving digital  

landscape — may exhibit distinct integration maturity patterns that differ  

from those documented in Western European or North American studies. The present  

thesis directly addresses this gap by providing the first empirical assessment  

of PMO IS integration maturity in Latvian enterprises. 

\begin{table}[h] 

  \caption{Comparison of related research strands} 

  \label{tab:related-work} 

  \centering 

  {\singlespacing 

  \begin{tabular}{p{0.25\linewidth}p{0.25\linewidth}p{0.22\linewidth}p{0.18\linewidth}} 

    \toprule 

    \textbf{Strand} & \textbf{Focus} & \textbf{Strength} & \textbf{Limitation}\\ 

    \midrule 

    Traditional PMO & Process standardization & Strong governance theory & No IS integration focus\\ 

    API Integration & Technical connectivity & Automation evidence & No PMO context\\ 

    BI \& Governance & Dashboard systems & Governance link & Large enterprises only\\ 

    Maturity Models & Capability levels & Structured framework & Not PMO IS specific\\ 

    Baltic Digital Maturity & Regional context & Latvia relevance & No PMO IS empirics\\ 

    \bottomrule 

  \end{tabular} 

  } 

\end{table} 

\section{Theoretical Framework} 

\label{sec:theory} 

  

This section presents the theoretical lens that grounds the empirical investigation.  

Based on the review of existing maturity models and IS integration research in  

Section~\ref{sec:related}, this thesis proposes a four-level Integration Maturity  

Model (IMM) specifically adapted to PMO information system environments. 

 

 

\subsection{Foundations of the Integration Maturity Model} 

The proposed model draws on three theoretical foundations. First, the staged  

progression concept from the Capability Maturity Model Integration (CMMI)  

\cite{paulk1993}, which posits that organizations advance through discrete,  

measurable levels of capability. Second, the enterprise integration patterns  

framework \cite{hohpe2004}, which classifies integration mechanisms by their  

degree of automation and architectural sophistication. Third, the data quality  

dimensions framework \cite{dama2017}, which links integration architecture to  

data accuracy, completeness, and timeliness outcomes. 

 

\subsection{The Four-Level Integration Maturity Model} 

The Integration Maturity Model proposed in this thesis defines four levels of  

IS integration maturity in PMO environments, as presented in  

Table~\ref{tab:imm-levels}. 

  

{\singlespacing 

\begin{table}[h] 

  \caption{Four-Level Integration Maturity Model for PMO Information Systems} 

  \label{tab:imm-levels} 

  \centering 

  \begin{tabular}{p{0.08\linewidth}p{0.15\linewidth}p{0.25\linewidth}p{0.22\linewidth}p{0.20\linewidth}} 

    \toprule 

    \textbf{Level} & \textbf{Name} & \textbf{Integration Mechanism} &  

    \textbf{Reporting} & \textbf{Governance}\\ 

    \midrule 

    1 & Initial & No integration; fully manual data consolidation via spreadsheets &  

    Fully manual; high error rate & Ad hoc; no standardization\\ 

    2 & Developing & File-based integration (CSV/Excel exports); partial automation &  

Semi-manual scheduled exports & Basic standards; inconsistent\\ 

    3 & Defined & API-based point-to-point integration; automated data flows &  

    Mostly automated; near real-time & Defined processes; mostly consistent\\ 

    4 & Optimized & Platform-level integration; unified data ecosystem with BI &  

    Fully automated; real-time dashboards & fully standardized continuous improvement\\ 

    \bottomrule 

  \end{tabular} 

\end{table} 

} 

\textbf{Level 1 --- Initial.}  

At this level, PMO environments depend on putting together data by manual means. 

 People usually use spreadsheets to put together reports from project data collected from many 

 different tools that aren't connected to each other. There is no organized way for data to move 

 between systems, which makes it very likely that mistakes, inconsistencies, and delays in 

 reporting will happen. There are no set rules for governance practices 

  

\textbf{Level 2 --- Developing.}  

At Level 2, some level of semi-automated, there is integration through file-based 

 data exchange, like when systems transmit CSV or Excel files to each other on a set schedule. 

 This helps to save some time, but people still need to do a lot of work to combine and check the 

 data. Reporting is partially automated, and governance standards are only partially defined and 

 not consistently followed. 

 

 

 

\textbf{Level 3 --- Defined.} Level 3 organizations have implemented API-based  

point-to-point integrations between key PMO systems, enabling automated data  

flows between project tracking, financial, and reporting tools. This enables data flow 

 automatically between tools for tracking projects, finances, and  

reports. Reporting is mostly done by computers and is almost real-time. Most PMO functions 

 have clear rules for governance that they always follow 

  

\textbf{Level 4 --- Optimized.} At the highest level of maturity, PMO environments  

operate within a unified platform ecosystem where all systems are fully integrated  

and data flows in real time. Centralized BI dashboards give you a full view of your portfolio 

, and all governance processes are fully standardized with ways to keep getting better.  

This level is the standard for how well information systems should work together in PMO 

 settings. 

 

\subsection{Variables and Relationships} 

The model defines the following key variables and relationships that will be  

tested empirically in Chapter~3: 

  

\begin{itemize} 

    \item \textbf{Independent variable:} Integration Maturity Level (1--4),  

    operationalized through survey items measuring integration mechanisms,  

    automation level, and BI centralization. 

    \item \textbf{Dependent variables:} Reporting automation degree, perceived  

    data accuracy and timeliness, and governance standardization level. 

    \item \textbf{Expected relationships:} Higher integration maturity levels  

    are expected to be positively associated with all three dependent variables,  

    consistent with the hypotheses stated in the Introduction. 

\end{itemize} 

\section{Research Motivation and Direction} 

\label{sec:motivation} 

 

\paragraph{Problem significance.} 

For Latvian enterprises, the fact that information systems get mixed up in PMO environments is 

a major operational problem. When project tracking, financial, and reporting tools don't work 

together, the PMO staff has to spend a lot of time manually putting together data, which raises 

the chance of mistakes and makes management reports less timely. This lack of efficiency has a 

 direct impact on how well strategic decisions are made at the portfolio level.   

The people who are most affected are PMO managers, project sponsors, and executive leaders 

 who need accurate and up-to-date project data to decide how to spend money and resources. 

 

\paragraph{Identified gap.} 

As established in Section~\ref{sec:related}, current research on IS integration maturity fails to 

 specifically focus on PMO environments, and there is a lack of empirical studies investigating 

 this phenomenon within the Latvian enterprise context. The gap can be stated  

precisely as follows: \emph{the distribution of IS integration maturity levels across  

PMO environments in Latvian enterprises, and its relationship with  

reporting automation, data quality, and governance standardization, has not been  

empirically measured or documented.} This gap is both testable and measurable  

through survey-based quantitative research. 

 

 

 

 

 

\paragraph{Direction and scope.} 

This thesis addresses the gap by conducting an empirical survey-based study of IS 

integration maturity in PMO environments in medium and large Latvian businesses.  

The research utilizes the four-level Integration Maturity Model established in Section 

 ~\ref{sec:theory} as the primary measurement framework.  

The scope is intentionally limited to enterprises with 50 or more employees that uphold a  

formal Project Management Office (PMO) structure, as smaller organizations are unlikely to 

 possess sufficiently intricate information systems landscapes to render integration maturity a 

 significant issue. The study explicitly excludes the analysis of individual project outcomes, 

 software implementation processes, or the technical architecture of specific integration 

 solutions. This direction is directly related to the goals and tasks set out in the introduction. 

 

\paragraph{Success criteria.} 

The study will be considered successful if it achieves the following outcomes:  

(1) a valid and reliable survey instrument is developed and administered to at  

least 30 respondents from Latvian enterprises with formal PMO structures;  

(2) the distribution of integration maturity levels across the sample is  

described using descriptive statistics; (3) statistically significant  

relationships between integration maturity and the dependent variables  

(reporting automation, data quality, governance standardization) are tested  

using correlation and regression analysis; and (4) practical recommendations  

for PMO IS integration improvement are formulated based on the empirical findings. 

\section{Chapter Summary} 

\label{sec:summary} 

  

This chapter established the theoretical and conceptual foundations of the thesis.  

Section~\ref{sec:concepts} defined eight key terms — PMO, PMIS, IS integration,  

integration maturity model, business intelligence, data quality, reporting  

automation, and medium/large enterprise — providing a consistent conceptual  

vocabulary for the remainder of the thesis. Section~\ref{sec:related} surveyed  

five thematic strands of related research, contrasting traditional PMO research,  

API integration studies, BI-governance research, maturity model frameworks, and  

Baltic digital maturity research, and identified a concrete empirical gap.  

Section~\ref{sec:theory} proposed a four-level Integration Maturity Model as  

the primary theoretical framework, defining the variables and relationships  

to be tested empirically. Section~\ref{sec:motivation} articulated the problem 

significance, the measurable research gap, the scope of the study, and the  

success criteria. 

  

\noindent\textbf{Conclusions for Chapter~1.} 

The literature review confirms that while substantial research exists on PMO  

governance, IS integration architectures, and digital maturity models, no  

empirical study has specifically examined IS integration maturity in PMO  

environments within Latvia. The proposed four-level Integration Maturity Model  

provides a theoretically grounded and operationalizable framework for addressing  

this gap. These foundations motivate the research methodology presented in  

Chapter~2 and the empirical study conducted in Chapter~3. 

 

 

 

 

  

 
