\chapter{Survey Questionnaire}
\label{chap:appendixA}

This appendix presents the complete survey instrument used to collect empirical
data for this study. The survey was distributed electronically via Google Forms
to representatives of medium and large Latvian enterprises with formal Project
Management Office (PMO) structures. Participation was voluntary and all
responses were collected anonymously.

\bigskip
\noindent\textbf{Survey title:} Assessment of Information System Integration
Maturity in Project Management Offices

\noindent\textbf{Estimated completion time:} 10--15 minutes

\noindent\textbf{Target respondents:} PMO managers, project managers, IT
managers, or digital transformation leads at organizations with 50 or more
employees and a formal PMO structure.

\bigskip

% ── PART A ──────────────────────────────────────────────────────────────────
\section*{Part A --- Organizational Profile}

\noindent\textbf{A1.} How many employees does your organization currently have?

\begin{itemize}
  \item[$\square$] 50--249 employees (medium-sized enterprise)
  \item[$\square$] 250--999 employees (large enterprise)
  \item[$\square$] 1000 or more employees (very large enterprise)
\end{itemize}

\bigskip
\noindent\textbf{A2.} What industry does your organization primarily operate in?

\begin{itemize}
  \item[$\square$] Information Technology
  \item[$\square$] Finance and Banking
  \item[$\square$] Manufacturing
  \item[$\square$] Retail and Commerce
  \item[$\square$] Telecommunications
  \item[$\square$] Public Sector
  \item[$\square$] Other (please specify): \underline{\hspace{5cm}}
\end{itemize}

\bigskip
\noindent\textbf{A3.} How long has your PMO been formally established?

\begin{itemize}
  \item[$\square$] Less than 1 year
  \item[$\square$] 1--3 years
  \item[$\square$] 3--5 years
  \item[$\square$] More than 5 years
\end{itemize}

\bigskip
\noindent\textbf{A4.} How many active projects does your PMO currently oversee?

\begin{itemize}
  \item[$\square$] 1--10 projects
  \item[$\square$] 11--25 projects
  \item[$\square$] 26--50 projects
  \item[$\square$] More than 50 projects
\end{itemize}

\bigskip

% ── PART B ──────────────────────────────────────────────────────────────────
\section*{Part B --- Integration Maturity Assessment}

\noindent\textbf{B1.} Which of the following best describes how data flows
between your PMO's information systems? \textit{(Select one option.)}

\begin{itemize}
  \item[$\square$] \textbf{Level 1 --- Initial:} Data is consolidated manually
  from multiple tools into spreadsheets. There are no automated connections
  between systems.
  \item[$\square$] \textbf{Level 2 --- Developing:} Some systems share data via
  scheduled file exports (e.g., CSV or Excel). Some manual work is still
  required to consolidate and verify data.
  \item[$\square$] \textbf{Level 3 --- Defined:} Systems are connected via APIs.
  Reports are generated automatically and data flows are mostly automated and
  near real-time.
  \item[$\square$] \textbf{Level 4 --- Optimized:} All systems are fully
  integrated within a unified platform with real-time dashboards and full
  portfolio visibility. Governance processes are fully standardized.
\end{itemize}

\bigskip
\noindent\textit{For items B2--B5, please indicate your level of agreement
using the following scale:}

\medskip
\begin{center}
\begin{tabular}{ccccc}
  \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5}\\
  Strongly Disagree & Disagree & Neutral & Agree & Strongly Agree
\end{tabular}
\end{center}

\bigskip

{\singlespacing
\begin{longtable}{@{} p{0.05\linewidth} p{0.65\linewidth} p{0.20\linewidth} @{}}
\toprule
\textbf{No.} & \textbf{Statement} & \textbf{Rating (1--5)}\\
\midrule
\endfirsthead
\toprule
\textbf{No.} & \textbf{Statement} & \textbf{Rating (1--5)}\\
\midrule
\endhead
\bottomrule
\endfoot
B2 & Our PMO systems share data automatically without manual intervention.
   & \underline{\hspace{2cm}}\\[6pt]
B3 & Our PMO uses API connections between project tracking, financial, and
   reporting tools.
   & \underline{\hspace{2cm}}\\[6pt]
B4 & Our PMO relies on manual data entry or file exports (CSV/Excel) to
   consolidate project data.
   & \underline{\hspace{2cm}}\\[6pt]
B5 & Our PMO operates within a unified platform ecosystem where all systems
   are fully integrated.
   & \underline{\hspace{2cm}}\\
\end{longtable}
}

% ── PART C ──────────────────────────────────────────────────────────────────
\section*{Part C --- Reporting Automation and Data Quality}

\noindent\textit{Please indicate your level of agreement with each statement
(1 = Strongly Disagree, 5 = Strongly Agree). Item C5 is reverse scored.}

\bigskip

{\singlespacing
\begin{longtable}{@{} p{0.05\linewidth} p{0.65\linewidth} p{0.20\linewidth} @{}}
\toprule
\textbf{No.} & \textbf{Statement} & \textbf{Rating (1--5)}\\
\midrule
\endfirsthead
\toprule
\textbf{No.} & \textbf{Statement} & \textbf{Rating (1--5)}\\
\midrule
\endhead
\bottomrule
\endfoot
C1 & Project status reports in our PMO are generated automatically from
   integrated systems.
   & \underline{\hspace{2cm}}\\[6pt]
C2 & Our PMO dashboards are updated in real time without manual intervention.
   & \underline{\hspace{2cm}}\\[6pt]
C3 & The data in our PMO reports is accurate and reliable.
   & \underline{\hspace{2cm}}\\[6pt]
C4 & Project data is available in a timely manner for decision-making.
   & \underline{\hspace{2cm}}\\[6pt]
C5 & We frequently encounter errors or inconsistencies in our PMO reporting
   data. \textit{(reverse scored)}
   & \underline{\hspace{2cm}}\\
\end{longtable}
}

% ── PART D ──────────────────────────────────────────────────────────────────
\section*{Part D --- Governance Standardization}

\noindent\textit{Please indicate your level of agreement with each statement
(1 = Strongly Disagree, 5 = Strongly Agree).}

\bigskip

{\singlespacing
\begin{longtable}{@{} p{0.05\linewidth} p{0.65\linewidth} p{0.20\linewidth} @{}}
\toprule
\textbf{No.} & \textbf{Statement} & \textbf{Rating (1--5)}\\
\midrule
\endfirsthead
\toprule
\textbf{No.} & \textbf{Statement} & \textbf{Rating (1--5)}\\
\midrule
\endhead
\bottomrule
\endfoot
D1 & Our PMO follows standardized processes for project reporting across all
   projects.
   & \underline{\hspace{2cm}}\\[6pt]
D2 & Our PMO uses centralized dashboards to monitor all projects in one place.
   & \underline{\hspace{2cm}}\\[6pt]
D3 & Project governance processes in our PMO are consistently applied across
   all departments.
   & \underline{\hspace{2cm}}\\[6pt]
D4 & Our PMO has clearly defined and documented standards for data collection
   and reporting.
   & \underline{\hspace{2cm}}\\[6pt]
D5 & The use of centralized BI tools has improved governance standardization
   in our PMO.
   & \underline{\hspace{2cm}}\\
\end{longtable}
}

\bigskip
\noindent\textit{Thank you for completing this survey. Your responses are
anonymous and will be used solely for academic research purposes in partial
fulfilment of the Bachelor's thesis requirements at Riga Nordic University.}
