\chapter{Research Methodology} 

\label{chap:analysis} 

 

This chapter presents the research methodology applied in this thesis. It describes the overall 

 research design, the operationalization of the Integration Maturity Model, the design of the 

 survey instrument and expert interviews, data collection procedure, and the statistical methods  

used to analyze the collected data.  

  

\section{Research Design and Approach} 

\label{sec:design} 

This thesis adopts a quantitative research design, complemented by qualitative expert 

 interviews. The principal research methodology is a cross-sectional survey study, in  

which the data is gathered from various organizations at a singular moment in time.  

This design is appropriate for the research objective, which is to evaluate the present distribution 

 of IS integration maturity levels within PMO environments in Latvian enterprises and to 

 examine its correlations with reporting automation, data quality, and governance 

 standardization \cite{saunders2019}. 

 

The research follows a deductive approach, in which the four hypotheses stated  

in the Introduction are derived from the theoretical framework presented in  

Chapter~1 and tested against empirical data collected from Latvian enterprises. 

  

\section{Operationalization of the Integration Maturity Model} 

\label{sec:operationalization} 

  

The four-level Integration Maturity Model (IMM) proposed in Chapter~1 serves as  

the primary measurement framework. To enable survey-based measurement, each  

maturity level is operationalized through a set of observable, measurable  

indicators, as presented in Table~\ref{tab:imm-operationalization}. 

  

{\singlespacing 

\begin{table}[h] 

  \caption{Operationalization of Integration Maturity Model levels} 

  \label{tab:imm-operationalization} 

  \centering 

  \begin{tabular}{p{0.08\linewidth}p{0.15\linewidth}p{0.35\linewidth}p{0.30\linewidth}} 

    \toprule 

    \textbf{Level} & \textbf{Name} & \textbf{Observable Indicators} &  

    \textbf{Survey Items}\\ 

    \midrule 

    1 & Initial & No system connections; all data consolidated manually in  

    spreadsheets; no automated reports &  

    "Data is consolidated manually from multiple tools into spreadsheets"\\ 

    2 & Developing & Some file-based exports between systems; partial  

    automation; scheduled report generation &  

    "Some systems share data via file exports (CSV/Excel)"\\ 

    3 & Defined & API connections between key systems; automated data flows;  

    near real-time reporting &  

    "Systems are connected via APIs; reports are generated automatically"\\ 

    4 & Optimized & Full platform integration; unified BI dashboard;  

    real-time portfolio visibility &  

    "All systems are fully integrated in a unified platform with real-time dashboards"\\ 

    \bottomrule 

  \end{tabular} 

\end{table} 

} 

Respondents are asked to select the description that best matches their  

organization's current PMO IS integration approach. The chosen level functions   

 as the primary independent variable in the study. 

  

\section{Survey Instrument Design} 

\label{sec:survey} 

  

The survey instrument was designed to find out the main ideas of the study:   

level of integration maturity, automation of reporting, quality of data, and standardization of 

 governance. There are four sections to the instrument. 

  

\begin{itemize} 

    \item \textbf{Part A --- Organizational Profile:} Questions about the  

    organization's size, industry, and PMO structure (categorical items). 

    \item \textbf{Part B --- Integration Maturity Assessment:} A single  

    categorical item in which respondents select the IMM level that best  

    describes their PMO's IS integration approach, supplemented by  

    Likert-scale items measuring specific integration practices. 

    \item \textbf{Part C --- Reporting Automation and Data Quality:}  

    Five-point Likert-scale items measuring the degree of reporting  

    automation and perceived data accuracy and timeliness. 

    \item \textbf{Part D --- Governance Standardization:} Five-point  

    Likert-scale items measuring the degree of governance standardization  

    in PMO processes. 

\end{itemize} 

  

All Likert-scale items use a five-point response scale ranging from  

1 (Strongly Disagree) to 5 (Strongly Agree). The survey was designed in  

English and distributed electronically via Google Forms. The estimated  

completion time is 10--15 minutes. 

  

\section{Survey Questionnaire} 

\label{sec:questionnaire} 

  

The complete survey instrument is presented in  

Table~\ref{tab:survey-questions}. Each item is aligned with a specific  

research construct and hypothesis. 

  

{\singlespacing 

\begin{table}[h] 

  \caption{Survey instrument items by construct and hypothesis} 

  \label{tab:survey-questions} 

  \centering 

  \begin{tabular}{p{0.05\linewidth}p{0.40\linewidth}p{0.25\linewidth}p{0.15\linewidth}} 

    \toprule 

    \textbf{No.} & \textbf{Survey Item} & \textbf{Construct} &  

    \textbf{Hypothesis}\\ 

    \midrule 

    \multicolumn{4}{l}{\textbf{Part A --- Organizational Profile}}\\ 

    A1 & How many employees does your organization have?  

    (50--249 / 250+) & Organization size & --\\ 

    A2 & What industry does your organization operate in?  

    (categorical) & Industry sector & --\\ 

    A3 & How long has your PMO been formally established?  

    (years) & PMO maturity & --\\ 

    A4 & How many active projects does your PMO currently oversee?  

    (numerical) & PMO scope & --\\ 

    \midrule 

    \multicolumn{4}{l}{\textbf{Part B --- Integration Maturity Assessment}}\\ 

    B1 & Which of the following best describes how data flows  

    between your PMO's information systems? (select one IMM level  

    description) & Integration maturity level & H1\\ 

    B2 & Our PMO systems share data automatically without manual  

    intervention. & Integration automation & H1, H2\\ 

    B3 & Our PMO uses API connections between project tracking,  

    financial, and reporting tools. & API integration & H2\\ 

    B4 & Our PMO relies on manual data entry or file exports  

    (CSV/Excel) to consolidate project data. & Manual integration & H1, H2\\ 

    B5 & Our PMO operates within a unified platform ecosystem  

    where all systems are fully integrated. & Platform integration & H1, H2\\ 

    \midrule 

    \multicolumn{4}{l}{\textbf{Part C --- Reporting Automation and Data Quality}}\\ 

    C1 & Project status reports in our PMO are generated  

    automatically from integrated systems. & Reporting automation & H2\\ 

    C2 & Our PMO dashboards are updated in real time without  

    manual intervention. & Reporting automation & H2\\ 

    C3 & The data in our PMO reports is accurate and reliable. &  

    Data accuracy & H3\\ 

    C4 & Project data is available in a timely manner for  

    decision-making. & Data timeliness & H3\\ 

    C5 & We frequently encounter errors or inconsistencies in  

    our PMO reporting data. (reverse scored) & Data quality & H3\\ 

    \midrule 

    \multicolumn{4}{l}{\textbf{Part D --- Governance Standardization}}\\ 

    D1 & Our PMO follows standardized processes for project  

    reporting across all projects. & Governance standardization & H4\\ 

    D2 & Our PMO uses centralized dashboards to monitor all  

    projects in one place. & BI centralization & H4\\ 

    D3 & Project governance processes in our PMO are consistently  

    applied across all departments. & Governance consistency & H4\\ 

    D4 & Our PMO has clearly defined and documented standards  

    for data collection and reporting. & Governance documentation & H4\\ 

    D5 & The use of centralized BI tools has improved governance  

    standardization in our PMO. & BI-governance link & H4\\ 

    \bottomrule 

  \end{tabular} 

\end{table} 

} 

  

\paragraph{Note on scaling.} 

All items in Parts B--D use a five-point Likert scale:  

1 = Strongly Disagree, 2 = Disagree, 3 = Neutral,  

4 = Agree, 5 = Strongly Agree. Item C5 is reverse scored,  

meaning higher values indicate lower data quality issues. 

  

\section{Expert Interview Design} 

\label{sec:interviews} 

  

To supplement the quantitative survey findings with qualitative depth,  

2--3 structured expert interviews are conducted with PMO leads, IT architects,  

or digital transformation managers from Latvian enterprises. The interviews  

follow a structured protocol with open-ended questions covering the following  

themes: 

  

\begin{itemize} 

    \item The current state of IS integration in the respondent's PMO environment 

    \item Challenges experienced due to low integration maturity 

    \item The role of BI and dashboard systems in governance standardization 

    \item Perceived barriers to advancing integration maturity 

    \item Practical recommendations for improving PMO IS integration 

\end{itemize} 

  

Interview responses are analyzed qualitatively using thematic analysis to  

identify patterns and insights that complement the quantitative findings. 

  

\section{Data Collection Procedure} 

\label{sec:datacollection} 

  

The survey is distributed to representatives of medium and large Latvian  

enterprises (50+ employees) with formal PMO structures. The target sample  

size is 30--50 responses, which is sufficient for the planned descriptive  

and correlational statistical analyses \cite{field2018}. Respondents are  

identified through professional networks, LinkedIn, and direct organizational  

contacts. Only one response per organization is accepted to ensure independence  

of observations. 

Data collection takes place over a period of four to six weeks. All responses  

are anonymous and stored securely in accordance with data protection principles.  

Participation is voluntary and respondents are informed of the purpose of the  

study prior to completing the survey. 

  

\section{Data Analysis Methods} 

\label{sec:analysis} 

  

The collected data is analyzed using the following statistical methods: 

  

\textbf{Descriptive statistics.} Frequency distributions and percentages are  

used to describe the distribution of integration maturity levels across the  

sample. Mean scores and standard deviations are calculated for Likert-scale  

items measuring reporting automation, data quality, and governance  

standardization. 

  

\textbf{Normality testing.} The Shapiro-Wilk test is applied to determine  

whether the data follows a normal distribution, which informs the choice  

between parametric (Pearson) and non-parametric (Spearman) correlation  

analysis \cite{field2018}. 

  

\textbf{Correlation analysis.} Pearson or Spearman correlation coefficients  

are calculated to examine the relationships between integration maturity level  

and the three dependent variables: reporting automation, data quality, and  

governance standardization. Statistical significance is assessed at the  

$p < 0.05$ level. 

  

\textbf{Regression analysis.} Simple linear regression is applied to examine  

the predictive relationship between integration maturity level and each  

dependent variable, allowing for quantification of the strength and direction  

of the relationships. 

  

\textbf{Qualitative analysis.} Expert interview transcripts are analyzed using  

thematic analysis to identify recurring themes and insights that supplement  

the quantitative findings. 

  

All quantitative analyses are conducted using SPSS or R statistical software. 

  

\section{Validity and Reliability} 

\label{sec:validity} 

  

To ensure the quality of the research instrument and findings, the following  

measures are taken to address validity and reliability. 

  

\textbf{Content validity.} The survey items were designed based on established  

theoretical frameworks — specifically the Integration Maturity Model proposed  

in Chapter~1, the CMMI framework \cite{paulk1993}, and the data quality  

dimensions from DAMA-DMBOK \cite{dama2017}. Each survey item is explicitly  

linked to a research construct and hypothesis, ensuring that the instrument  

measures what it is intended to measure. 

  

\textbf{Construct validity.} The survey constructs — integration maturity,  

reporting automation, data quality, and governance standardization — are  

operationalized based on definitions from peer-reviewed literature and  

established frameworks. Multiple items per construct are included to reduce  

measurement error and improve construct coverage. 

  

\textbf{Internal consistency reliability.} Cronbach's alpha coefficient will  

be calculated for each multi-item construct scale to assess internal  

consistency. A Cronbach's alpha value of 0.70 or above is considered  

acceptable for research purposes \cite{field2018}. 

  

\textbf{External validity.} The study targets a specific population —  

representatives of medium and large Latvian enterprises with formal PMO  

structures. While the findings may not be directly generalizable to all  

organizations globally, they are intended to be representative of the  

Latvian enterprise context within the defined scope. 

  

\section{Research Ethics} 

\label{sec:ethics} 

  

This study adheres to the following ethical principles throughout the data  

collection and analysis process: 

  

\begin{itemize} 

    \item \textbf{Voluntary participation.} All survey respondents and  

    interview participants participate voluntarily and may withdraw at  

    any time without consequence. 

    \item \textbf{Informed consent.} Respondents are informed of the  

    purpose of the study, how their data will be used, and who will  

    have access to the results before completing the survey. 

    \item \textbf{Anonymity and confidentiality.} Survey responses are  

    collected anonymously. No personally identifiable information is  

    collected or stored. Interview participants are anonymized in the  

    thesis unless explicit consent for identification is provided. 

    \item \textbf{Data protection.} All collected data is stored securely  

    and processed in accordance with the General Data Protection Regulation  

    (GDPR) and applicable Latvian data protection legislation. 

    \item \textbf{Honest reporting.} All findings are reported accurately  

    and without manipulation, including results that do not support the  

    research hypotheses. 

\end{itemize} 

  

\section{Limitations of the Research Design} 

\label{sec:limitations} 

  

While this study is designed to provide a rigorous empirical assessment of  

IS integration maturity in PMO environments in Latvia, several limitations  

must be acknowledged. 

  

\textbf{Self-reported data.} The survey relies on self-reported perceptions  

of PMO representatives, which may be subject to social desirability bias —  

respondents may rate their organization's integration maturity higher than  

it actually is. To mitigate this, the survey includes specific behavioral  

indicators rather than general self-assessment questions. 

  

\textbf{Sample size.} The target sample of 30--50 responses is sufficient  

for descriptive and correlational analysis but limits the statistical power  

of regression models and prevents subgroup analysis by industry or  

organization size. Future research with larger samples would strengthen  

the generalizability of findings. 

  

\textbf{Cross-sectional design.} The study captures a snapshot of integration  

maturity at a single point in time and cannot establish causal relationships  

between integration maturity and the dependent variables. Longitudinal  

research would be needed to confirm causal direction. 

  

\textbf{Geographic scope.} The study is deliberately bounded to Latvian  

enterprises, which limits the generalizability of findings to other national  

or regional contexts. However, this scope is appropriate given the research  

aim of providing Latvia-specific empirical evidence. 

  

\textbf{Construct measurement.} Integration maturity is measured through  

a self-developed instrument rather than a previously validated scale, as no  

existing validated scale specifically measures IS integration maturity in  

PMO environments. While the instrument is grounded in established theoretical  

frameworks, its psychometric properties require further validation in  

future research. 

  

\section{Chapter Summary} 

\label{sec:chap2summary} 

  

This chapter described the research methodology applied in this thesis.  

A quantitative cross-sectional survey design complemented by qualitative expert  

interviews was adopted. The four-level Integration Maturity Model was  

turned into measurable survey indicators that could be used in real life. 

The survey instrument was designed to measure integration maturity, reporting automation, 

 data quality, and governance standardization across PMO environments in Latvian enterprises.  

Data collection targets 30--50 responses from medium and large Latvian  

enterprises with formal PMO structures. Statistical analysis methods including  

descriptive statistics, normality testing, correlation analysis, and regression  

analysis are applied to test the four research hypotheses. Validity, reliability,  

ethical principles, and research limitations were also addressed. The results  

of the empirical study are presented in Chapter~3.  

 

 
